---
title: "Reproducible Research Course Project 1"
author: "Chris Schmidt"
date: "08/12/2020"
output:
  pdf_document: default
  html_document: default
---


### Step 1 - Read in data ###

To commence the data analysis, following forking of project from github, it is necessary
to load the relevant R packages for transformation and analysis.  The data must then be extracted
to allow the by unzipping the files and reading them into an R object.
In this project I utilise the data.table package, with fread being the function used
to read .csv files to an R object.

```{r Preliminary, include=FALSE}
library(tidyverse)
library(data.table)
library(ggplot2)
library(xtable)
```

```{r Step 1 Code}
unzip("./activity.zip", exdir = "./data")
data <- fread(input = "./data//activity.csv")
```


### Step 2 - Histogram of total number of steps taken each day ###

To complete this step, I have first calculated the steps for each data using data.table,
and stored this as hist_data.  ggplot2 was then used with a geom_col and appropriate
titles to display the required histogram.

```{r Step 2 Code, fig.height=4}
daily_data <- data[, .(`daily_steps` = sum(steps, na.rm = TRUE)), by = date]

plot1 <- ggplot(data = daily_data, mapping = aes(x = date, y = daily_steps))
plot1 + geom_col() +
        ggtitle("Total Daily Steps", ) + xlab("Date") + ylab("Total Steps") +
        theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))
```


### Step 3 - Mean and median of number of steps taken each day ###

```{r Step 3 Code}
daily_data[, .(`Mean Daily Steps` = mean(daily_steps, na.rm = TRUE),
         `Median Daily Steps` = median(daily_steps, na.rm = TRUE))]
```


### Step 4 - Time series plot of the average number of steps taken ###

To calculate these results it is necessary to process the data so that it is grouped
by time interval, which has been achieved using the `by = x` syntax in the data.table
package.  This data has then been plotted using the ggplot 2 package and the geom_line()
function.

```{r Step 4 Code, fig.height=5}
interval_data <- data[, .(`interval_average` = mean(steps, na.rm = TRUE)), by = interval]
plot2 <- ggplot(data = interval_data, mapping = aes(x=interval, y = interval_average))
plot2 + geom_line() +
        ggtitle("Mean Steps by Time Interval") + xlab("Time") + ylab("Mean Steps") +
        theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))
```


### Step 5 - Calculate the five minute interval that on average contains the maximum number of steps ###

This is a simple case of using the which.max function to caculate the row of the interval_data object
which has the largest number of average steps.  This has then been returned in a statement by using the
paste function.

```{r Step 5 Code}
max_interval <- which.max(interval_data[, `interval_average`])
paste ("5 minute interval with most average steps is", interval_data[max_interval, interval])
```


### Step 6 - Code to describe a strategy for imputing missing data ###

In order to find the number of rows with NA in the steps column, we simply use the function
is.na on the relevant vector and wrap it in the sum function.

```{r Step 6a Code}
nasteps <- sum(is.na(data$steps))
```

This result tells us that there are `r nasteps` NAs in the steps column in the raw dataset.

Missing data (NAs) has been imputed with the interval mean, which has already been
calculated in the interval_data object, and updating the steps in the data table
with this mean value.  I note some days also have "0" values for steps, but no attempt
is made to impute this data, as legitimate "0" values for steps also appear in the data,
and repairing the data in this way is beyond the scope of the assignment.

This has been achieved by joining the interval_data DT to the data DT using data.table's
`on = ` syntax.  The result has been stored in a new DT imputed_data to clearly identify
that it is no longer the raw data which is being shown in the steps column.

The following line of code identifies where an NA value exists in the steps column,
an updates it with the (rounded) interval average (as steps is an integer value).

This work has been check by calculating by testing is.na for the imputed_data$steps vector,
which returns 0 (ie. TRUE on 0 rows).

```{r Step 6b Code}
imputed_data <- data[interval_data, on = "interval"]
imputed_data[is.na(steps), steps := round(interval_average,0)]
sum(is.na(imputed_data$steps))
```


### Step 7 - Histogram of the total number of steps taken each day after missing values imputed ###

This is completed using the imputed_data object and following the same strategy outlined
for the original histogram generated using the raw data only.

```{r Step 7 Code}
imputed_daily_data <- imputed_data[, .(daily_steps = sum(steps)), by = date]
plot3 <- ggplot(data = imputed_daily_data, mapping = aes(x = date, y = daily_steps))
plot3 + geom_col() +
        labs(title = "Total Steps by Date", subtitle = "NB: Missing values have been imputed") +
        xlab("Date") + ylab("Total Daily Steps") +
        theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
              plot.subtitle = element_text(face = "italic", size = 14, hjust = 0.5))
```


### Step 8 - Comparison of weekday and weekend activity levels

To complete this step the imputed_data object has been transformed by adding a factor variable
denoting each observation as belonging to either "Weekday" or "Weekend".  This was achieved
through use of an intermediate step of adding a day variable for each observation, derived
from the date.

Mean steps for each discrete interval were then calculated and grouped according to whether
the observation were of weekdays or weekend days.

Finally this data is plotted using the ggplot package, with geom_line and by constructing
separate panels for weekdays and weekend days using the facet_wrap function.

```{r Step 8 Code, fig.height=8}
imputed_data[, `:=` (day = weekdays(date), weekday_flag = as.factor("Weekday"))]
imputed_data[day %in% c("Saturday", "Sunday"), weekday_flag := as.factor("Weekend")]
imputed_interval_data <- imputed_data[, .(interval_average = mean(steps)), by = .(interval, weekday_flag)]
plot4 <- ggplot(data = imputed_interval_data, mapping = aes(x=interval, y = interval_average))
plot4 + geom_line() + facet_wrap(.~weekday_flag, nrow = 2) +
        labs(title = "Average Steps by Time Interval", subtitle = "NB: Missing values have been imputed") +
        xlab("Interval (Time of Day)") + ylab("Average Steps") +
        theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
              plot.subtitle = element_text(face = "italic", size = 14, hjust = 0.5))
```


